---
title: "Chapter 07: Exploratory Graphs"
output: html_document
author: Nutsa Nanuashvili
---

```{r}
suppressMessages(library(tidyverse))
```

Import data Air pollution (average PM2.5)
```{r}
path = "C:/Users/nutsa/Documents/Projects/EDA_PaiR/data/"
polut_data = readr::read_csv( paste0 (path, "avgpm25.csv"), 
                 col_types = cols( pm25 = col_double(),
  fips = col_character(),
  region = col_factor(),
  longitude = col_double(),
  latitude = col_double() ) )
```
Check data
```{r}
glimpse(polut_data)
```
```{r}
levels(polut_data$region)

slice_sample(polut_data, n = 10)
```
### Features

* `fips` - 5-digit code indicating the county
* `pm25` - average level
* `region` - region of the country in which the county resides
* `longitude` and `latitude` of the centroid for that county

```{r}
summary(polut_data)
```

```{r}
boxplot(polut_data$pm25, col = "green" )
```

Drow a line at national standard of 12 micrograms
```{r}
boxplot(polut_data$pm25, col = "green")
abline(h = 12)
```

Look at the points above pm25 = 15, because the standard must not exeed 12 micrograms
```{r}
polut_data %>%
  filter(pm25 > 15)
```

```{r}
suppressMessages(library(maps))

maps::map("county", "california")
with(filter(polut_data, pm25 >15), points (longitude, latitude, col = "red" ))
```


### Histogram
```{r}
hist(polut_data$pm25, col = "blue", breaks = 100)
rug(polut_data$pm25)
```

Draw two lines, one at the median of the data and one at 12, the level of the standard
```{r}
hist(polut_data$pm25, col = "blue")
rug(polut_data$pm25)
abline(v = median(polut_data$pm25), col = "magenta", lwd = 4)
abline(v = 12, lwd = 3)
```


### Barplot
Illustrate the number of counties in each region - west and east
```{r}

region = polut_data$region
table(region) %>% barplot(col = "grey", main = "County Count for Each Region")
```


Difference in PM2.5 levels between the eastern and western parts of the U.S.

```{r}
boxplot(pm25 ~ region, data = polut_data, col = "wheat", main = "PM25 per region of the country")
abline(h = 12, col = "gray")
```


Plotting 2 histograms of PM25 for east and west regions separately

```{r}
# dividing data into regions
east = polut_data %>% filter(region =="east") 
west = polut_data %>% filter (region == "west") 

# set up the margins and location of the plot
par(mfrow = c(2,1), mar = c(3, 3, 1, 1))
# plot data for each region
hist(east$pm25, col = "blue", main = "PM25 level in east counties")
hist(west$pm25, col = "blue", main = "PM25 level in west counties")

```


### Scatterplots

`with` evaluates _“an R expression in an environment constructed from data”_. 

```{r}
# plot pm25 as a function of latitude per region
with(polut_data, plot(latitude, pm25, col = polut_data$region))
#lty defines the line style
abline(h = 12, col = "blue", lty = 2)
```

To see which region mapped on which color we have to look at factor levels because 1st level `col = 1` is black (the default color) and `col = 2` is red.

Compare east and west
```{r}
par(mfrow = c(1, 2), mar = c(5, 4, 2, 1))

plot(west$latitude, west$pm25, main = "West")
abline(h = 12, col = "blue",lwd = 2,  lty = 3)
plot(east$latitude, east$pm25, main = "East")
abline(h = 12, col = "blue", lwd = 2, lty = 3)
```









